Self-tunable approximated explicit MPC: Heat exchanger implementation and analysis

### 2023-02-03 paper/elsarticle/main.tex
FINALLY

	%% TODO: UNCOMMENT THIS SECTION TO COMPILE ALSO INITIAL PART OF THE PAPER
	%
	%\begin{frontmatter}
		%
		%%% Title, authors and addresses
		%
		%%% use the tnoteref command within \title for footnotes;
		%%% use the tnotetext command for theassociated footnote;
		%%% use the fnref command within \author or \address for footnotes;
		%%% use the fntext command for theassociated footnote;
		%%% use the corref command within \author for corresponding author footnotes;
		%%% use the cortext command for theassociated footnote;
		%%% use the ead command for the email address,
		%%% and the form \ead[url] for the home page:
		%%% \title{Title\tnoteref{label1}}
		%%% \tnotetext[label1]{}
		%%% \author{Name\corref{cor1}\fnref{label2}}
		%%% \ead{email address}
		%%% \ead[url]{home page}
		%%% \fntext[label2]{}
		%%% \cortext[cor1]{}
		%%% \affiliation{organization={},
			%%%             addressline={},
			%%%             city={},
			%%%             postcode={},
			%%%             state={},
			%%%             country={}}
		%%% \fntext[label3]{}
		%
		%\title{Self-tunable approximated explicit MPC: Heat exchanger implementation and analysis}
		%%\title{Self-tunable approximated explicit MPC: Implementation on a heat exchanger with an asymmetric behavior}
		%
		%%% use optional labels to link authors explicitly to addresses:
		%%% \author[label1,label2]{}
		%%% \affiliation[label1]{organization={},
			%%%             addressline={},
			%%%             city={},
			%%%             postcode={},
			%%%             state={},
			%%%             country={}}
		%%%
		%%% \affiliation[label2]{organization={},
			%%%             addressline={},
			%%%             city={},
			%%%             postcode={},
			%%%             state={},
			%%%             country={}}
		%
		%\author{Lenka Galčíková}
		%\author{Petronela Belková}
		%\author{Juraj Oravec}
		%
		%\affiliation{organization={Slovak University of Technology in Bratislava, Faculty of Chemical and Food Technology, Institute of Information Engineering, Automation, and Mathematics,},%Department and Organization
			%            addressline={Radlinského 9}, 
			%            city={Bratislava},
			%            postcode={81237}, 
			%%            state={Slovakia},
			%            country={Slovakia}
			%}
		%
		%\begin{abstract}
			%%max 250 words!

			The tunable approximated explicit model predictive control (MPC) comes with the benefits of real-time tunability without the necessity of solving the optimization problem online. This paper provides a novel self-tunable control policy that does not require any further tuning or interventions of the control engineer during operation. Based on the current operating conditions, the autonomous tuning parameter scales the control input using linear interpolation between the boundary optimal control actions. The adjustment of the tuning parameter depends on the current reference value, which makes this strategy suitable for reference tracking problems. Furthermore, a novel technique for scaling the tuning parameter is proposed. It provides to exploit different ranges of the tuning parameter assigned to specified operating conditions. The self-tunable explicit MPC was implemented on a laboratory heat exchanger with nonlinear and asymmetric behavior. To ensure offset-free reference tracking control, the built-in integrator was included in the explicit MPC design. The asymmetric behavior of the plant was compensated by tuning the controller's aggressiveness, as the negative or positive sign of reference change was considered in the tuning process. The designed self-tunable controller improved control performance by decreasing integral squared error, maximal overshoots/undershoots and settling time compared to the conventional control strategy based on a single (not tuned) controller. 
			%\end{abstract}
		%
		%%%Graphical abstract
		%\begin{graphicalabstract}
			%%\includegraphics{grabs}
			%\end{graphicalabstract}
		%
		%%%Research highlights
		%\begin{highlights}
			%% 3-5 highlights, every one should not exceed 85 characters including spaces
			%\item Self-tunable control technique does not need any intervention during control
			%\item Tuning based on the operating conditions compensates for the nonlinear plant behavior
			%\item Implementation of the proposed tuning method on a laboratory heat exchanger 
			%\item Control performance improved compared to utilizing only one explicit MPC
			%\end{highlights}
		%
		%\begin{keyword}
			%tunable explicit MPC \sep self-tunable technique \sep tuning parameter \sep heat exchanger \sep control performance
			%%% keywords here, in the form: keyword \sep keyword
			%
			%%% PACS codes here, in the form: \PACS code \sep code
			%
			%%% MSC codes here, in the form: \MSC code \sep code
			%%% or \MSC[2008] code \sep code (2000 is the default)
			%
			%\end{keyword}
		%
		%\end{frontmatter}

@@

	The heat exchangers in their numerous variants are integrated into the many industrial plants as the heat transfer represents the crucial phenomena for all thermal energy applications~\cite{KV18}. Simultaneously, the heat exchangers are energy-demanding and resource-intensive plants. Moreover, from the control viewpoint, the controller design taking into account the nonlinear and asymmetric behavior of the heat exchangers is a challenging task, see~\cite{RL20}. Although the conventional and widely-used proportional–integral–derivative (PID) controllers are robust and easy to tune, their control performance may not be sufficient when it comes to energy savings. Various extensions built above the well-tuned PID controller were developed to compensate for the nonlinear and asymmetric behavior, often affected by the additional impact of the uncertain parameters originated, e.g., in the fouling~\cite{MT19}. Such widely-used control strategies including, e.g., the robust control~\cite{WY18}, the gain-scheduling, self-tuning, adaptive and optimal control, see~\cite{MF08} for the rigorous mathematical modeling and controller design methods in general, and~\cite{Liptak} for the controller design tailored for the process control engineers.

	One of the promising control strategies addressing all these issues in an optimal way came with the formation and tuning of the model predictive control (MPC), e.g., see~\cite{Morari_MPC}. MPC provides optimal control input based on the minimization of a specified cost function while considering a model of the system. Compared to linear quadratic controllers (LQR)~\cite{LQR}, model predictive control also includes constraints on the control input or process variables~\cite{Maciejowski_MPC}, and additional saturation is not necessary. Moreover, as the optimization problem is solved in every control step, MPC represents a receding horizon control policy~\cite{receding_horizon}, having a significant benefit mainly in the terms of disturbance rejection. The model predictive control was intensively investigated in connection with heat exchangers. In~\cite{Vinaya_HE_MPC}, the authors developed a model predictive control for a shell and tube heat exchanger. In~\cite{Oravec_energy}, the control performance of the robust MPC implemented for a plate heat exchanger with soft constraints was analyzed. Four robust control strategies were presented and compared in~\cite{Oravec_HE_ATE}. A two-level control structure was applied on a heat exchanger network in~\cite{Gonzales_HE_MPC}, where the low level of control was ensured by MPC and the high level by a supervisory online optimizer. The fast nonlinear MPC was designed to optimize the waste heat recovery~\cite{WC19}.  The multi-layer control designed in~\cite{DZ18} designed the MPC in the leader loop to optimize the thermal response to improved control performance.

@@

	The plant on which the control was implemented and analyzed is a laboratory-scaled counter-current liquid-to-liquid plate heat exchanger Armfield Process Plant Trainer PCT23~\cite{pct23}, see Figure~\ref{fig:HE}. The heat exchanger is 90\,mm wide, 103\,mm long, and 160\,mm high. The heat exchange is performed between the cold feed (water) and the hot medium (water). The cold feed as well as the heating medium are transported to the heat exchanger by two peristaltic pumps with flexible tubing from silicon rubber. The flow rate of the feed is constant, while the aim of control is to track the reference value of the feed temperature. Therefore, the controlled variable is the feed temperature $T$ at the outlet of the heat exchanger. The temperature of the heated feed in the outlet stream was measured by the type K thermocouple. The associated manipulated variable is the voltage $U$ corresponding to the power of the pump feeding the heat exchanger by the hot medium. The voltage is within the range of [$0-5$]\,V normalized into the relative values in percentage. For further technical specifications of the laboratory heat exchanger, the reader is referred to~\cite{pct23}. As heat exchange is a nonlinear and asymmetric process~\cite{Liptak}, this heat exchanger represents a suitable candidate for the presented controller tuning strategy. The corresponding illustrative scheme of the implemented closed-loop control setup is in Figure~\ref{fig:CL}, where the ``self-tuning'' block substitutes the more detailed scheme in Figure~\ref{fig:tuning}.

@@

	\begin{figure}
		\begin{center}
			\includegraphics[width=\textwidth]{images/cl}
			\caption[Scheme of the implemented closed-loop control setup.]{Closed-loop scheme.}
			\label{fig:CL}
		\end{center}
	\end{figure}

@@

	To analyze the control from the environmental viewpoint, the energy consumption was also investigated. Technically, the total power supply utilized to heat the heating medium was measured and summarized for the whole control but for each controller setup, respectively. The total energy consumption $E$ consumed by each controller as well as the relative improvement subject to each boundary controller is summarized in Table~\ref{tab:energy}.

@@

	It can be seen in Table~\ref{tab:energy}, that utilizing the self-tuned controller leads to energy consumption which acquires value between the energy consumption corresponding to the boundary controllers. Although it is not the minimum one, the energy is still saved compared to the boundary controller with $Q_\mathrm{y} = 100$. Note, though the controller with $Q_\mathrm{y} = 1\,000$ leads to minimum energy consumption, using this single non-tunable controller does not ensured the required control performance. As for the test bed served the laboratory-scaled heat exchanger plant, the greater energy savings are expected for the industrial-scaled plants.

@@

	This paper deals with the experimental implementation and analysis of the novel self-tunable approximated explicit model predictive control method and provides a strategy for an effective self-tuning controller design. Based on the current value of the piece-wise constant reference, the tuning parameter is scaled using linear interpolation. 
	In contrast to the previously published work related to the self-tunable explicit MPC suggested tuning based on the distance of the reference value from the system steady-state value. This paper presents a novel perspective idea of the self-tuning based on the size of reference step change. The self-tuning algorithm aims to compensate for the nonlinear behavior of the controlled system. The self-tuning parameter is updated whenever the reference changes. the tuning value is calculated as the ratio between the size of the reference change and the maximal admissible size of the reference change, which is specified before operation. 
	Another novel contribution addresses the challenging control problem of asymmetric system behavior by splitting the interval of the self-tuning parameter into two ranges, while both intervals are assigned to different operating conditions. The proposed method is implemented on a laboratory-scaled heat exchanger with nonlinear and asymmetric behavior. The asymmetry makes the plant a suitable candidate to analyze the benefits of splitting the interval of the tuning parameter. The decision criterium is negativity or positivity of reference change. When the reference changed upwards, the control input was tuned in the first part of the interval and approached the boundary controller associated with the lower bound on the selected penalty matrix. On the contrary, when the reference changed downwards, the control input was tuned to approach the control input from the boundary controller with the upper bound on the penalty matrix. To investigate the control results properly, the control performance was judged also quantitatively using a set of quality criteria. Compared to the self-tunable control approach outperformed the conventional control strategy handling just a single controller, i.e., non-tunable controller, by minimizing the integral squared control error up to 32\,\%, reducing maximal overshoots/undershoots up to 18\,\%, and decreasing settling time up to 24\,\%. In addition, the energy consumption corresponding to each of the constructed controllers was also analyzed. The energy consumption of the self-tunable explicit MPC controller was in between the values of the two boundary non-tunable explicit MPC controllers, i.e., the total energy consumption trades off the performance of the boundary controllers. % did not deteriorate in general. 
	With saving the energy compared to one of the boundary controllers, the control performance was significantly improved.

>>>>

	%% TODO: UNCOMMENT THIS SECTION TO COMPILE ALSO INITIAL PART OF THE PAPER
	%
	\begin{frontmatter}

		%% Title, authors and addresses

		%% use the tnoteref command within \title for footnotes;
		%% use the tnotetext command for theassociated footnote;
		%% use the fnref command within \author or \address for footnotes;
		%% use the fntext command for theassociated footnote;
		%% use the corref command within \author for corresponding author footnotes;
		%% use the cortext command for theassociated footnote;
		%% use the ead command for the email address,
		%% and the form \ead[url] for the home page:
		%% \title{Title\tnoteref{label1}}
		%% \tnotetext[label1]{}
		%% \author{Name\corref{cor1}\fnref{label2}}
		%% \ead{email address}
		%% \ead[url]{home page}
		%% \fntext[label2]{}
		%% \cortext[cor1]{}
		%% \affiliation{organization={},
			%%             addressline={},
			%%             city={},
			%%             postcode={},
			%%             state={},
			%%             country={}}
		%% \fntext[label3]{}

		\title{Self-tunable approximated explicit MPC: Heat exchanger implementation and analysis}
		%\title{Self-tunable approximated explicit MPC: Implementation on a heat exchanger with an asymmetric behavior}

		%% use optional labels to link authors explicitly to addresses:
		%% \author[label1,label2]{}
		%% \affiliation[label1]{organization={},
			%%             addressline={},
			%%             city={},
			%%             postcode={},
			%%             state={},
			%%             country={}}
		%%
		%% \affiliation[label2]{organization={},
			%%             addressline={},
			%%             city={},
			%%             postcode={},
			%%             state={},
			%%             country={}}

		\author{Lenka Galčíková}
		\author{Petronela Belková}
		\author{Juraj Oravec}

		\affiliation{organization={Slovak University of Technology in Bratislava, Faculty of Chemical and Food Technology, Institute of Information Engineering, Automation, and Mathematics,},%Department and Organization
			            addressline={Radlinského 9}, 
			            city={Bratislava},
			            postcode={81237}, 
			%            state={Slovakia},
			            country={Slovakia}
			}

		\begin{abstract}
			%%max 250 words!

			The tunable approximated explicit model predictive control (MPC) comes with the benefits of real-time tunability without the necessity of solving the optimization problem online. This paper provides a novel self-tunable control policy that does not require any further tuning or interventions of the control engineer during operation. Based on the current operating conditions, the autonomous tuning parameter scales the control input using linear interpolation between the boundary optimal control actions. The adjustment of the tuning parameter depends on the current reference value, which makes this strategy suitable for reference tracking problems. Furthermore, a novel technique for scaling the tuning parameter is proposed. It provides to exploit different ranges of the tuning parameter assigned to specified operating conditions. The self-tunable explicit MPC was implemented on a laboratory heat exchanger with nonlinear and asymmetric behavior. To ensure offset-free reference tracking control, the built-in integrator was included in the explicit MPC design. The asymmetric behavior of the plant was compensated by tuning the controller's aggressiveness, as the negative or positive sign of reference change was considered in the tuning process. The designed self-tunable controller improved control performance by decreasing integral squared error, maximal overshoots/undershoots and settling time compared to the conventional control strategy based on a single (not tuned) controller. 
			\end{abstract}
		%
		%%%Graphical abstract
		%\begin{graphicalabstract}
			%%\includegraphics{grabs}
			%\end{graphicalabstract}
		%
		%%Research highlights
		\begin{highlights}
			% 3-5 highlights, every one should not exceed 85 characters including spaces
			\item Self-tunable control technique does not need any intervention during control
			\item Tuning based on operating conditions compensates for the nonlinear plant behavior
			\item Implementation of the proposed tuning method on a laboratory heat exchanger 
			\item Control performance improved compared to utilizing only one explicit MPC
			\end{highlights}

		\begin{keyword}
			tunable explicit MPC \sep self-tunable technique \sep tuning parameter \sep heat exchanger \sep control performance
			%% keywords here, in the form: keyword \sep keyword

			%% PACS codes here, in the form: \PACS code \sep code

			%% MSC codes here, in the form: \MSC code \sep code
			%% or \MSC[2008] code \sep code (2000 is the default)

			\end{keyword}

		\end{frontmatter}

@@

	The heat exchangers in their numerous variants are integrated into the many industrial plants as the heat transfer represents the crucial phenomena for all thermal energy applications~\cite{KV18}. Simultaneously, the heat exchangers are energy-demanding and resource-intensive plants. Moreover, from the control viewpoint, the controller design taking into account the nonlinear and asymmetric behavior of the heat exchangers is a challenging task, see~\cite{RL20}. Although the conventional and widely-used proportional–integral–derivative (PID) controllers are robust and easy to tune, their control performance may not be sufficient when it comes to energy savings. Various extensions built above the well-tuned PID controller were developed to compensate for the nonlinear and asymmetric behavior, often affected by the additional impact of the uncertain parameters originated, e.g., in the fouling~\cite{MT19}. Such widely-used
	control strategies include, e.g., the robust control~\cite{WY18}, the gain-scheduling, adaptive and optimal control. For the rigorous mathematical modeling and controller design methods in general, see~\cite{MF08} and for the controller design tailored for the process control engineers see~\cite{Liptak}.

	One of the promising control strategies addressing all these issues in an optimal way came with the formation of the model predictive control (MPC), e.g., see~\cite{Morari_MPC}. MPC provides optimal control input based on the minimization of a specified cost function while considering a model of the system. Compared to linear quadratic controllers (LQR)~\cite{LQR}, model predictive control also includes constraints on the control input or process variables~\cite{Maciejowski_MPC}, and additional saturation is not necessary. Moreover, as the optimization problem is solved in every control step, MPC represents a receding horizon control policy~\cite{receding_horizon}, having a significant benefit mainly in the terms of disturbance rejection. The model predictive control was intensively investigated in connection with heat exchangers. In~\cite{Vinaya_HE_MPC}, the authors developed a model predictive control for a shell and tube heat exchanger. In~\cite{Oravec_energy}, the control performance of the robust MPC implemented for a plate heat exchanger with soft constraints was analyzed. Four robust control strategies were presented and compared in~\cite{Oravec_HE_ATE}. A two-level control structure was applied on a heat exchanger network in~\cite{Gonzales_HE_MPC}, where the low level of control was ensured by MPC and the high level by a supervisory online optimizer. The fast nonlinear MPC was designed to optimize the waste heat recovery~\cite{WC19}.  The multi-layer control designed in~\cite{DZ18} designed the MPC in the leader loop to optimize the thermal response to improved control performance.

@@

The plant on which the control was implemented and analyzed is a laboratory-scaled counter-current liquid-to-liquid plate heat exchanger Armfield Process Plant Trainer PCT23~\cite{pct23}, see Figure~\ref{fig:HE}. The heat exchanger is 90\,mm wide, 103\,mm long, and 160\,mm high. The heat exchange is performed between the cold feed (water) and the hot medium (water). The cold feed as well as the heating medium are transported to the heat exchanger by two peristaltic pumps with flexible tubing from silicon rubber. The flow rate of the feed is constant, while the aim of control is to track the reference value of the feed temperature. Therefore, the controlled variable is the feed temperature $T$ at the outlet of the heat exchanger. The temperature of the heated feed in the outlet stream was measured by the type K thermocouple. The associated manipulated variable is the voltage $U$ corresponding to the power of the pump feeding the heat exchanger by the hot medium. The voltage is within the range of [$0-5$]\,V normalized into the relative values in percentage. For further technical specifications of the laboratory heat exchanger, the reader is referred to~\cite{pct23}. As heat exchange is a nonlinear and asymmetric process~\cite{Liptak}, this heat exchanger represents a suitable candidate for the presented controller tuning strategy. The corresponding illustrative scheme of the implemented closed-loop control setup is in Figure~\ref{fig:CL}, where the ``Self-tuning'' block substitutes the more detailed scheme in Figure~\ref{fig:tuning}.

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=\textwidth]{images/cl}
		\caption[Scheme of the implemented closed-loop control setup.]{Scheme of the implemented closed-loop control setup.}
		\label{fig:CL}
	\end{center}
\end{figure}

@@

To analyze the control from the environmental viewpoint, energy consumption was also investigated. Technically, the total power supply utilized to heat the heating medium was measured and summarized for the whole control but for each controller setup respectively. The total energy consumption $E$ consumed by each controller as well as the relative improvement subject to each boundary controller is summarized in Table~\ref{tab:energy}.

@@

It can be seen in Table~\ref{tab:energy}, that utilizing the self-tuned controller leads to energy consumption which acquires value between the energy consumption corresponding to the boundary controllers. Although it is not the minimum one, the energy is still saved compared to the boundary controller with $Q_\mathrm{y} = 100$. Note, though the controller with $Q_\mathrm{y} = 1\,000$ leads to minimum energy consumption, using this single non-tunable controller does not ensure the required control performance. As the laboratory-scaled heat exchanger plant served as the testbed, greater energy savings are expected for the industrial-scaled plants.

@@

		This paper deals with the experimental implementation and analysis of the novel self-tunable approximated explicit model predictive control method and provides a strategy for an effective self-tuning controller design. Based on the current value of the piece-wise constant reference, the tuning parameter is scaled using linear interpolation. 
	The previously published work related to the self-tunable explicit MPC suggested tuning based on the distance of the reference value from the system steady-state value. This paper presents a novel perspective idea of self-tuning based on the size of reference step change. The self-tuning algorithm aims to compensate for the nonlinear behavior of the controlled system. The self-tuning parameter is updated whenever the reference changes. The tuning value is calculated as the ratio between the size of the reference change and the maximal admissible size of the reference change, which is specified before operation. 
	Another novel contribution addresses the challenging control problem of asymmetric system behavior by splitting the interval of the self-tuning parameter into two ranges, while both intervals are assigned to different operating conditions. The proposed method is implemented on a laboratory-scaled heat exchanger with nonlinear and asymmetric behavior. The asymmetry makes the plant a suitable candidate to analyze the benefits of splitting the interval of the tuning parameter. The decision criterium is negativity or positivity of reference change. When the reference changed upwards, the control input was tuned in the first part of the interval and approached the boundary controller associated with the lower bound on the selected penalty matrix. On the contrary, when the reference changed downwards, the control input was tuned to approach the control input from the boundary controller with the upper bound on the penalty matrix. To investigate the control results properly, the control performance was judged also quantitatively using a set of quality criteria. The self-tunable control approach outperformed the conventional control strategy handling just a single controller, i.e., non-tunable controller, by minimizing the integral squared control error up to 32\,\%, reducing maximal overshoots/undershoots up to 18\,\%, and decreasing settling time up to 24\,\%. In addition, the energy consumption corresponding to each of the constructed controllers was also analyzed. The energy consumption of the self-tunable explicit MPC controller was in between the values of the two boundary non-tunable explicit MPC controllers. % the total energy consumption trades off the performance of the boundary controllers. % did not deteriorate in general. 
	To conclude, with saving energy compared to one of the boundary controllers, the control performance was significantly improved.
